Fail Faster

### 2025-03-12 paper/paper.tex
some notes from meeting

@@

@@

\subsection{Multi-Stage Programming}

@@

\section{Why is \bq\ slow, and how can we fix it?}

@@

@@

The crux of our binding time analysis is that particular size random seed are
only known at run time (the later stage), but the code of the generator itself
is known at compile time.

Generators --- values of type \texttt{'a Bq.t} --- are never constructed dynamically at run time in practice,
so the

@@

\jwc{Note that some of this is a lie, but it's a good first cut: point to the section on \emph{The Trick} to see the real version.}

\jwc{
  \begin{itemize}
    \item The simplest combinators are presented below.
    \item \texttt{return} is the constant generator, given code for \texttt{x : 'a}, it gives the generator which always generates \texttt{x}
    \item \texttt{bind} is like BQ's bind --- sequencing generators by passing the result of one to another --- except that the continuation gets access to the
    \texttt{code} for the result of running the first generator. At compile time, we know only that \texttt{g} will return \emph{some} \texttt{a}, but we do not get access to the particular \texttt{a}.
    \item Bind takes code for the size and seed, and returns code that first let-binds the result of the first generator.
    The function application \texttt{g size\_c random\_c} runs during the first stage, producing the code for the \texttt{a}. This is then spliced into the quote: the specialized code of the generator \texttt{g}
    put on the right-hand side of the let. Then, we splice in the code for the continuation, passing first the variable x, and the same size and random seed.
  \end{itemize}
}

\begin{lstlisting}
module Gen = struct
  type 'a t = int code -> Random.t code -> 'a code

@@

  let bind (g : 'a t) (k : 'a code -> 'b t) : 'b t =
    fun size_c random_c ->
      .<
        let x = .~(g size_c random_c) in
        .~(k .<x>. size_c random_c)
      >.

@@

  let to_bq (g : 'a code Gen.t) : ('a Bq.t) code =
  .<
    fun size random -> .~(g .<size>. .<random>.)
  >.
end
\end{lstlisting}

\jwc{Conversion to a bq generator}

\jwc{Then show the running example, with }

@@

>>>>

\jwc{People have been using staged metaprogrammign to eliminate abstraction
overheads in parsing for a long time, we turn that lens on generation (Which looks like parsing)}

@@

\jwc{
Let's lead on with the quote from PBT in practice: performance engineering
matters.  De-emphasize the abstraction overhead framing. Eliminating abstraction
overhead is important, so is using the fastest RNG you can possibly use.
}

@@

\subsection{Multi-Stage Programming}
\label{subsection:msp}

@@

\section{Why is \bq\ slow, and how can we fix it?}
\jwc{Why is a monadic generator library slow?}

@@

\jwc{If you're not using the fasteest RNG possible, you're leaving RNG on the table.}

@@

The crux of our binding time analysis is that the particular random seed and
size parameter \jwc{if explained} are only known at run time (the later stage),
but the code of the generator itself is known at compile time. Generators ---
values of type \texttt{'a Bq.t} --- are never constructed dynamically at run
time in practice, so all of the combinators we use to build them can run at compile time.

@@

This type, along with basic monadic generator DSL functionality can be found in Figure~\ref{fig:gen-staged-basic}.
The monadic interface is given by a return and bind, as usual.
\texttt{return} is the constant generator, but this time it operates on codes. Given \texttt{cx : 'a code}, the code for a \texttt{'a}, it returns the generator which always generates that value.
\texttt{bind g k} sequences generators by passing the result of running the generator \texttt{g} to a continuation \texttt{k}. However, instead of getting access
to the particular value generated by \texttt{g}, the continuation \texttt{k} gets access to \texttt{code} for the value sampled from \texttt{g}:
at compile time, we only know \texttt{g} will generate \emph{some} \texttt{'a}, but not which one.
Operationally, bind takes code for the size and seed, and returns code that (1) let-binds a variable \texttt{a} to spliced-in code that runs \texttt{g}, and then
(2) runs the spliced-in continuation \texttt{k}.
Both function applications \texttt{g size\_c random\_c}
and \texttt{k .<a>. size\_c random\_c} run at compile time.
\texttt{Gen.int} is the generator that samples an int from the RNG.
Given any size and random seed, it returns a code block that calls \texttt{SR.int} with that random seed. Because the lower and upper bounds
might not be known at compile time --- they may themselves be the results of calling \texttt{Gen.int} ---
the arguments \texttt{lo} and \texttt{hi} are of type \texttt{int code}, and get spliced into the code block as arguments to \texttt{SR.int}.
Lastly, \texttt{to_bq} turns a staged generator into code for a normal \bq
generator. This function is just a 2-argument version of ``The Trick'' (\texttt{eta}
from Section~\ref{subsection:msp}).

\begin{figure}
\begin{lstlisting}
module Gen = struct
  type 'a t = int code -> Random.t code -> 'a code

@@

  let bind (g : 'a t) (k : 'a code -> 'b t) : 'b t =
    fun size_c random_c ->
      .<
        let a = .~(g size_c random_c) in
        .~(k .<a>. size_c random_c)
      >.

@@

  let to_bq (g : 'a code Gen.t) : ('a Bq.t) code =
  .<
    fun size random -> .~(g .<size>. .<random>.)
  >.
end
\end{lstlisting}
\caption{Basic Staged Generator Library}
\ref{fig:gen-staged-basic}
\end{figure}

Returning to our running example, we can define a staged generator for pairs of ints

@@

\jwc{Changing some identifier names }
